The Evolution of Phenotypically Invariant Genetic Networks

### 2017-08-04 kalman_walk_paper.tex
rewrote quant gen section

% fitness as a fn of distance
\newcommand{\fit}{\mathcal{F}}
% fitness as a fn of A
\newcommand{\fitx}{\mathcal{f}}
% set of optimal coefficients
\newcommand{\optx}{\mathcal{X}}
% optimal phenotype
\newcommand{\optph}{\Phi_0}
% distance in phenotype space
\newcommand{\dph}{d}

@@

Since the argument applies more generally, we abstract the model above.
Suppose that a set $x$ of coefficients that determine a system (this is $A$ above),
produce a phenotype $\Phi(x)$ (the time course of $\phi(t)$).
There is an optimal phenotype $\optph$,
and a set $\optx$ of ``optimal'' coefficients that produce this phenotype.
Fitness depends on distance to the optimal phenotype --
we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \fit(\dph(\Phi(x),\optph)/\beta)$,
where $\beta$ controls the scale on which fitness drops.
We will assume that the map $\Phi$ and the distance function $\dph()$ are smooth,
and that the optimal set $\optx$ is locally isomorphic to $\R^m$.
\plr{say that better}

\paragraph{Offspring.}
Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
and the individual's coefficients are the average of her two haploid values (no dominance).
This implies that the diploid population variance, $\sigma^2$, is one-half the haploid variance.
Each new gamete is produced from the parent's two haploid copies;
for simplicity we assume that the gamete inherits a random choice of one of the two parental copies,
and so $\sigma$ remains constant, up to a $1/N_e$ term.
A more general model including segregation variance \citep{barton_infinitesimal}
would result in the same conclusions.
\plri{but put this in the appendix?}

\paragraph{System drift}
% plr how fast does it drift
If the variation within a population of some coefficient
has standard deviation $\sigma$,
then since subsequent generations resample from this diversity,
the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
simply because this is the the standard deviation of the mean of a random sample \citep{lande_drift}.
% (This could be taken as a definition of $N_e$.)
Selection will tend to restrain this motion,
but mean movement along the optimal set $\optx$ is unconstrained.
% (although perhaps complicated by recombination load \citep{recomb_load})
The value of $\sigma$ in particular directions in coefficient space 
depends on constraints imposed by selection and 
the covariance between genetic variation between different coefficients -- known as the $G$ matrix \citep{G_matrix}.
% covariance which may arise due to functional constraints and/or statistical linkage.
% There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
For instance, linkage will lead to covariance
if the variation is due to \textit{cis}-regulatory variants,
in which case the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
across which a population may carry only a few common haplotypes.
However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
(leading to positive covariance between $A_{ik}$ and $A_{jk}$)
is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
(Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)

@@

To obtain a general quantitative picture, we need to know 
$\sigma_N$ and $\sigma_S$, the standard deviations of coefficient variation along and perpendicular to $\optx$ respectively;
$\gamma$, how fast phenotype changes moving away from $\optx$;
and $\beta$, how fast fitness drops with distance from the optimal phenotype.
(Concretely, $\gamma$ is the derivative of $\dph(\Phi(x+u z), \Phi(x))$ with respect to $u$ for $x \in \optx$ and $z$ perpendicular to the tangent subspace at $x$.)
With these parameters, a typical individual will have a fitness of around $\fit(\sigma_S \gamma / \beta)$.

\paragraph{Hybridization}
\plr{I put F1s and F2s here, but with less explanation than below. Merge somehow.}
By the arguments above, 
over $T$ generations the population mean will move a distance of order $\sqrt{\sigma_N/T}$ along $\optx$.
A $F_1$ hybrid between the two populations
has one haploid genome from each,
so will be distributed as the mean of two randomly chosen haploids from the two populations,
and so will have mean at the midpoint between their means,
and variance equal to $\sigma$.
A $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
as before,
but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations.
These are depicted in figure \ref{fig:conceptual_fig}.

\begin{figure}
\centering
\includegraphics{figures/conceptual_fig}
\caption{
    \label{fig:conceptual_fig}
    WRITE CAPTION
    \plri{The figure is not quite right now -- should show all types of F2 in purple.  but then it will be messy.  ideas?}
}
\end{figure}

\plri{improve figure by putting labels on from the following}
\plr{I think we can set $\beta=1$ by defining $\fit$ appropriately?}
Suppose that two populations have drifted independently to differ by $z$,
and that $z$ is of the same order as $\sigma$ but is smaller than $\gamma$,
so that for $x_0$ a point nearby on $\optx$,
\begin{align*}
    \Phi(x_0 + x) = \Phi(x_0) + D \Phi(x_0) \cdot x + O(\|\gamma x\|^2),
\end{align*}
and so $d(\Phi(x_0+x), \Phi(x_0)) \approx  \gamma x$.
The mean $F_1$ is the average of the parental means,
and since the first-order terms in the Taylor series vanish,
has phenotype differing from the optimum by a distance of order $(\gamma z)^2$.
\plri{(Proof: let the parental means be $x_1$ and $x_2$; then as $z=x_2-x_1$,
    $d(\Phi((x_1+x_2)/2), \Phi(x_1)) = D \Phi(x_1) \cdot z/2 + O(\gamma z)^2$,
    but expanding about $x_2$ instead we get
    $d(\Phi((x_1+x_2)/2), \Phi(x_2)) = - D \Phi(x_2) \cdot z/2 + O(\gamma \|z\|)^2$.
    Since $\Phi(x_1) = \Phi(x_2)$ and $D \Phi(x_1) - D \Phi(x_2) = O(\|z\|^2)$,
    we see that $D \Phi(x_1) \cdot z = O(\gamma z)^2$.}
The mean $F_2$ is the same,
but is a distance away from the optimal set of order $z$.
In summary, the typical fitness of an individual in the original population is
$\fit_0 = \fit(\sigma_s \gamma / \beta)$;
of an $F_1$ is
$\fit_1 = \fit((\sigma_s \gamma + \gamma z^2) / \beta)$;
and of an $F_2$ is
$\fit_2 = \fit(\gamma z / \beta)$.

If $\fit(x) = e^{-x^2} \approx 1 - x^2$,
then ...
\plri{reparameterize to define $\fit_1$ as the \emph{drop} in fitness?}

@@

The fitness of an organism can be computed by comparing its impulse response with the optimal response,
\plr{this is either zero or infinite.  apply a weighting function (see appendix)?}
\begin{align}
    \mathcal{F}\left( \widehat{\phi}(t) \right) = \exp \left\{- \frac{1}{\sigma} \int_{0}^{\infty} \left\lVert \phi(t) - \widehat{\phi}(t) \right\rVert^{2} dt  \right\}  .
\end{align}
Therefore a hybrid's fitness can be computed by comparing its impulse response with that of its parents.

@@

We quantitatively describe the degree of incompatibility between two populations $P_{1}$ and $P_{2}$ as
\begin{align*}
      \mathcal{I} = \frac{2 \left\langle \mathcal{F} \left(\phi_{F_{1}}\right) \right\rangle}{\left\langle \mathcal{F} \left(\phi_{P_{1}}\right) \right\rangle +  \left\langle \mathcal{F} \left(\phi_{P_{2}}\right) \right\rangle} ,
\end{align*}
where angled brackets imply averaging.  \plr{averaging over what? And, this is as others define - cite.}

\plr{This is not part of the definition of $\mathcal{I}$:}
F1s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
\plr{don't need these to be big equations.}
\begin{align*}
  \mathcal{F} \left(\phi_{P_{1}}\right) = \mathcal{F} \left( \phi_{P_{2}} \right) &= 1 \\
  \mathcal{F}\left(\phi_{F_{1}}\right) &= 0 \\
  \mathcal{I} &= 0
\end{align*}

@@

In general terms, if fitness $\mathcal{F}$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
\begin{align*}
  \mathcal{F} = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
\end{align*}
\plr{sentences don't begin with "And"}
And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intra-population regulatory variance,
\plr{what is $T$?}
\begin{align*}
  P_{T} \sim \mathcal{N} \left(A(\tau), T \frac{\sigma^{2}}{N_{e}}\right)
\end{align*}
the population mean architecture will move at rate
\plr{what is $\Delta \tau$? Also, if we reframe in terms of $V$ not $\tau$ above will need to reframe this.}
\begin{align*}
  \Delta \tau = \sigma \sqrt{\frac{T}{N_{e}}}
\end{align*}
where $T$ is time measured in generations. 
USING GENERAL MATH FROM MULTIVARIATE TAYLOR SERIES.... we know that the phenotype of a first generation $F_{1}$ hybrid diverges from that of its parents (assuming both parents are at an optimum and phenotypically equivalent) quartically as a function of $\Delta \tau$. 
Thus the squared difference of it's phenotype will be,
\begin{align*}
  \Delta \Phi^{2}_{1} = c_{1}^{2} \sigma^{4} \left( \frac{T}{N_{e}} \right)^{2}
\end{align*}
and $F_{1}$ fitness will decline linearly with respect to $\frac{T}{N_{e}}$. ARGUMENT FOR WHY $\sigma^{2} c_{1} \sim \beta$ HERE.
   \begin{align*}
     \Delta \mathcal{F}_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
\end{align*}
 It follows FROM THE SAME MATH ABOUT THE MULTIVAR TAYLOR that the phenotypes of second generation $F_{2}$ hybrid crosses diverges quadratically as a function of $\Delta \tau$, 
\begin{align*}
  \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
  \Delta \mathcal{F}_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
\end{align*}
and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE.

@@

\section{Differentiating the fitness function}

>>>>

% fitness as a fn of distance
\newcommand{\fit}{\mathcal{F}}
% fitness as a fn of A
\newcommand{\fitx}{\mathcal{F}}
% set of optimal coefficients
\newcommand{\optx}{\mathcal{X}}
% optimal phenotype
\newcommand{\optph}{\Phi_0}
% distance in phenotype space
\newcommand{\dph}{d}
% incompatibility
\newcommand{\Incompat}{\mathcal{I}}

@@

Suppose that a set $x$ of coefficients that determine a system (this is $A$ above),
produce a phenotype $\Phi(x)$ (the time course of $\phi(t)$).
There is an optimal phenotype $\optph$,
and a set $\optx$ of ``optimal'' coefficients that produce this phenotype.
Fitness depends on distance to the optimal phenotype --
we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
measured so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \exp(-\dph(\Phi(x),\optph)^2)$.
We will assume that the map $\Phi$ is smooth
and that the optimal set $\optx$ is locally isomorphic to $\R^m$.
\plr{say that better}

\paragraph{Offspring.}
Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
and the individual's coefficients are the average of her two haploid values (no dominance).
This implies that the diploid population variance, $\sigma^2$, is one-half the haploid variance.
Each new gamete is produced from the parent's two haploid copies;
for simplicity we assume that the gamete inherits a random choice of one of the two parental copies,
and so $\sigma$ remains constant, up to a $1/N_e$ term.
A more general model including segregation variance \citep{barton_infinitesimal}
would result in the same qualitative conclusions.
\plri{but put this in the appendix?}

\paragraph{System drift}
% plr how fast does it drift
If the variation within a population of some coefficient
has standard deviation $\sigma$,
then since subsequent generations resample from this diversity,
the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
simply because this is the the standard deviation of the mean of a random sample \citep{lande_drift}.
% (This could be taken as a definition of $N_e$.)
Selection will tend to restrain this motion,
but mean movement along the optimal set $\optx$ is unconstrained.
% (although perhaps complicated by recombination load \citep{recomb_load})
The amount of variance in particular directions in coefficient space 
depend on constraints imposed by selection and 
the covariance between genetic variation between different coefficients (the $G$ matrix \citep{G_matrix}).
% covariance which may arise due to functional constraints and/or statistical linkage.
% There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
For instance,
if the variation is due to \textit{cis}-regulatory variants,
the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
across which a population may carry only a few common haplotypes.
However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
(leading to positive covariance between $A_{ik}$ and $A_{jk}$)
is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
(Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)

@@

To obtain a general quantitative picture, we need to know 
$\sigma_N$ and $\sigma_S$, the standard deviations of coefficient variation along and perpendicular to $\optx$ respectively,
and $\gamma$, the scale on which phenotype changes moving away from $\optx$.
Concretely, $\gamma$ is the inverse of the derivative of $\dph(\Phi(x+u z), \Phi(x))$ 
with respect to $u$ for $x \in \optx$ and $z$ perpendicular to the tangent space at $x$.
With these parameters, a typical individual will have a fitness of around $\exp(-(\sigma_S/\gamma)^2)$.

\paragraph{Hybridization}
\plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.}
By the arguments above, the means of two allopatric populations separated for $T$ generations
will be a distance of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means,
and variance equal to $\sigma$.
Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
as before,
but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations.
These are depicted in figure \ref{fig:conceptual_fig}.

\begin{figure}
\centering
\includegraphics{figures/conceptual_fig}
\caption{
    \label{fig:conceptual_fig}
    A conceptual figure of the fitness consequences of hybridization:
    axes represent system coefficients (i.e., entries of $A$);
    the line of optimal system coefficients is down in black;
    dotted lines give phenotypic distances to the optimum.
    Two pairs of parental populations are shown in black, along the optimum;
    a hypothetical population of $F_1$s are shown for each in red,
    and the distribution of one type of $F_2$ is shown in purple
    (other types of $F_2$ are not shown).
    Solid lines depict the distance of the $F_2$ to optimum.
    \plri{Should show all types of F2? would be messy.}
}
\end{figure}

\plri{improve figure by putting labels on from the following}
Suppose that two populations have drifted independently to differ by $z$,
and that $z$ is of the same order as $\sigma$ but is smaller than $\gamma$.
The mean $F_1$ is the average of the parental means,
and since the first-order terms in the Taylor series vanish,
has phenotype differing from the optimum by a distance of order $\|z\|^2$ 
(see appendix \ref{apx:away_from_opt}).
The mean $F_2$ is the same,
but the standard deviation is of order $z$,
so that up to lower order terms, 
while the typical fitness of an individual in the original population is
$\fit_0 = \exp(-(\sigma_s/\gamma)^2)$;
of an $F_1$ is
$\fit_1/\fit_0 = \exp(-(c_1 \sigma_N^2 T/N_e)^2)$;
and of an $F_2$ is
$\fit_2/\fit_0 = \exp(- T/(N_e\gamma^2))$.

@@

The fitness of an organism can be computed by comparing its impulse response with the optimal response,
\plr{this is either zero or infinite.  apply a weighting function (see appendix)?}
\begin{align}
    \fit\left( \widehat{\phi}(t) \right) = \exp \left\{- \frac{1}{\sigma} \int_{0}^{\infty} \left\lVert \phi(t) - \widehat{\phi}(t) \right\rVert^{2} dt  \right\}  .
\end{align}
Therefore a hybrid's fitness can be computed by comparing its impulse response with that of its parents.

@@

We quantitatively describe the degree of incompatibility between two populations $P_{1}$ and $P_{2}$ as
\begin{align*}
      \Incompat = \frac{2 \left\langle \fit \left(\phi_{F_{1}}\right) \right\rangle}{\left\langle \fit \left(\phi_{P_{1}}\right) \right\rangle +  \left\langle \fit \left(\phi_{P_{2}}\right) \right\rangle} ,
\end{align*}
where angled brackets imply averaging.  \plr{averaging over what? And, this is as others define - cite.}

\plr{This is not part of the definition of $\mathcal{I}$:}
F1s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
\plr{don't need these to be big equations.}
\begin{align*}
  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
  \fit\left(\phi_{F_{1}}\right) &= 0 \\
  \mathcal{I} &= 0
\end{align*}

@@

In general terms, if fitness $\fit$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
\begin{align*}
  \fit = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
\end{align*}
\plr{sentences don't begin with "And"}
And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intra-population regulatory variance,
\plr{what is $T$?}
\begin{align*}
  P_{T} \sim \mathcal{N} \left(A(\tau), T \frac{\sigma^{2}}{N_{e}}\right)
\end{align*}
the population mean architecture will move at rate
\plr{what is $\Delta \tau$? Also, if we reframe in terms of $V$ not $\tau$ above will need to reframe this.}
\begin{align*}
  \Delta \tau = \sigma \sqrt{\frac{T}{N_{e}}}
\end{align*}
where $T$ is time measured in generations. 
USING GENERAL MATH FROM MULTIVARIATE TAYLOR SERIES.... we know that the phenotype of a first generation $F_{1}$ hybrid diverges from that of its parents (assuming both parents are at an optimum and phenotypically equivalent) quartically as a function of $\Delta \tau$. 
Thus the squared difference of it's phenotype will be,
\begin{align*}
  \Delta \Phi^{2}_{1} = c_{1}^{2} \sigma^{4} \left( \frac{T}{N_{e}} \right)^{2}
\end{align*}
and $F_{1}$ fitness will decline linearly with respect to $\frac{T}{N_{e}}$. ARGUMENT FOR WHY $\sigma^{2} c_{1} \sim \beta$ HERE.
   \begin{align*}
     \Delta \fit_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
\end{align*}
 It follows FROM THE SAME MATH ABOUT THE MULTIVAR TAYLOR that the phenotypes of second generation $F_{2}$ hybrid crosses diverges quadratically as a function of $\Delta \tau$, 
\begin{align*}
  \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
  \Delta \fit_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
\end{align*}
and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE.

@@

\section{Away from the optimum}
\label{apx:away_from_opt}

Let two points  on $\optx$ be $x_1$ and $x_2$, let $\bar x = (x_1+x_2)/2$, and let $z=(x_2 - x_1)/2$.
Then with $D \Phi$ and $D^2 \Phi$ the first and second derivatives of $\Phi$, respectively,
then Taylor expanding about $x_1$ and $x_2$ finds that
\begin{align*}
    \Phi(\bar x) 
    &= \Phi(x_1) + D\Phi(x_1) \cdot z + \frac{1}{2} z^T D^2 \Phi(x_1) z + O(\|z\|^3) \\
    &= \Phi(x_2) - D\Phi(x_2) \cdot z + \frac{1}{2} z^T D^2 \Phi(x_2) z + O(\|z\|^3) .
\end{align*}
Now, since $\Phi(x_1) = \Phi(x_2) = \optph$ and
\begin{align*}
    D\Phi(x_2) = D\Phi(x_1) + 2 z^T D^2 \Phi(x_1) + O(\|z\|^2), \quad \text{and} \\
    D^2\Phi(x_2) = D^2\Phi(x_1) + O(\|z\|), \quad \text{and} \\
\end{align*}
adding together the two equations above and dividing by two gets that
\begin{align*}
    \Phi(\bar x) 
    &= \optph - \frac{3}{2} z^T D^2 \Phi(x_1) z + O(\|z\|^3) .
\end{align*}

\section{Differentiating the fitness function}
\label{apx:H_calc}
